[Normalization] A pipeline to analyse time-course gene expression data

### 2020-10-07 scripts/manuscript.Rmd
FIX loading the float latex package
Preparing for submission

keywords: >
    time-course gene expression data, clustering, differential expression,
    workflow
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output:
    BiocWorkflowTools::f1000_article: default
    ms_document: default
bibliography: bibliography.bib
always_allow_html: yes
---

@@

Before we dive into the exploratory analysis and quality control, let us
define color schemes for our data that we will use across the whole analysis.
We define color schemes for groups and time points as named vectors. We also
define a series of markers (or plotting symbols) to distinguish replicate
samples in scatter plots. We also reorder the factor `Group` which describes the treatments so that the treatments are order from low to high pathogeny (with Control being first).

@@

In both
situations, we first need to set up a object (a `Moanin` object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, ...). The `Moanin` object
will contain information used
throughout this analysis, in particular the condition and timepoints of each sample and the
the basis matrix for fitting splines models.

We start by creating the `Moanin` object using the `create_moanin_model`
function. We need to provide two things to the function: a `data.frame` with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata `data.frame` object should contain at least
two columns: one named `Group`, containing the  treatement effect, and a
second one named `Timepoint` containing the timepoint information.

@@

We then use the `plot_splines_data`  function, only now applied to the centroids, to visualize the centroids of
each cluster obtained with the splines k-means model.

```{r fig.height=6, fig.width=6, fig.cap="K-means centroids"}
plot_splines_data(de_moanin_model,
    data=kmeans_clusters$centroids, 
    colors=ann_colors$Group,
    smooth=TRUE)
```

@@

### Assigning genes to clusters
As we can see, while these genes have some similarity with the pattern of the cluster centroids, these particularly genes are not the best examples of the cluster, in the sense of matching the centroid estimates. Because of this variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.

@@

- **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
  the elbow methods looks at the total within cluster sum of squares 
  as a function of the number of clusters (WCSS). When adding clusters
  doesn't decrease the WCSS by a sufficient
  amount, one can consider stopping. This method thus provides visual aid to
  the user to choose the number of clusters, but often the "elbow" is hard to
  see on real data, where the number of clusters is not clearly defined.
- **Silhouette method**. Similarly to the Elbow method, the Silhouette method
  refers to a method of validation of consistency within clusters, and
  provides visual aid to choose the number of clusters.
- **Stability methods** Stability methods are more computationally intensive
  than any other method, as they rely on assessing the stability of the
  clustering for every $k$ to a small randomization of the data. The user is
  then invited to choose the number of cluster based on a number of similarity
  measures.

First, let us run the clustering for all values of $k$ of interest. We
will, for each clustering, conserve (1) with within cluster sum of
squares; (2) the clustering assignement (or label) for each gene.

@@

Each column correspond to a bootstrap sample, each row to a gene, and each
entry to the label found for that particular clustering. Thus, for each gene,
we have an assignment to a cluster over the 20 resampling runs. For example,
for $k=5$:

```{r,echo=FALSE}
pander(head(stability_5))
```

>>>>

keywords: >
    time-course gene expression data, clustering, differential expression,
    workflow
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
header-includes:
   - \usepackage{float}
output: rticles::arxiv_article
bibliography: bibliography.bib
always_allow_html: yes
---

<!--

Change output format to
BiocWorkflowTools::f1000_article: default  for F1000Research layout
ms_document: default for word document

-->

@@

Before we dive into the exploratory analysis and quality control, let us
define color schemes for our data that we will use across the whole analysis.
We define color schemes for groups and time points as named vectors. We also
define a series of markers (or plotting symbols) to distinguish replicate
samples in scatter plots. We also reorder the factor `Group` which describes
the treatments so that the treatments are ordered from low to high pathogeny
(with Control being first).

@@

In both situations, we first need to set up an object (a `Moanin` object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, ...). The `Moanin`
object will contain information used throughout this analysis, in particular
the condition and timepoints of each sample and the basis matrix for
fitting splines models.

We start by creating the `Moanin` object using the `create_moanin_model`
function. We need to provide two things to the function: a `data.frame` with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata `data.frame` object should contain at least
two columns: one named `Group`, containing the  treatment effect, and a
second one named `Timepoint` containing the timepoint information.

@@

We then use the `plot_splines_data`  function, only now applied to the
centroids, to visualize the centroids of each cluster obtained with the
splines k-means model (Figure \ref{fig:centroids}).

```{r centroids, fig.height=6, fig.width=6, fig.cap="K-means centroids"}
plot_splines_data(de_moanin_model,
    data=kmeans_clusters$centroids, 
    colors=ann_colors$Group,
    smooth=TRUE)
```

@@

### Assigning genes to clusters

As we can see, while these genes have some similarity with the pattern of the
cluster centroids, these particular genes are not the best examples of the
cluster, in the sense of matching the centroid estimates. Because of this
variability in how well the genes fit a cluster, we would like to be able to
score how well a gene fits a cluster.

@@

- **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
  the elbow method looks at the total within cluster sum of squares 
  as a function of the number of clusters (WCSS). When adding clusters
  doesn't decrease the WCSS by a sufficient
  amount, one can consider stopping. This method thus provides visual aid to
  the user to choose the number of clusters, but often the "elbow" is hard to
  see on real data, where the number of clusters is not clearly defined.
- **Silhouette method**. Similarly to the Elbow method, the Silhouette method
  refers to a method of validation of consistency within clusters, and
  provides visual aid to choose the number of clusters.
- **Stability methods** Stability methods are more computationally intensive
  than any other method, as they rely on assessing the stability of the
  clustering for every $k$ to a small randomization of the data. The user is
  then invited to choose the number of clusters based on a number of similarity
  measures.

First, let us run the clustering for all values of $k$ of interest. We
will, for each clustering, conserve (1) with within cluster sum of
squares; (2) the clustering assignment (or label) for each gene.

@@

Each column corresponds to a bootstrap sample, each row to a gene, and each
entry to the label found for that particular clustering. Thus, for each gene,
we have an assignment to a cluster over the 25 resampling runs. For example,
for the first 10 bootstrap samples and $k=5$:

```{r,echo=FALSE}
pander(head(stability_5[, 1:10]))
```

### 2020-10-07 scripts/miscellaneous/normalization.Rmd
Preparing for submission

---
title: "[Normalization] A pipeline to analyse time-course gene expression data"
author:
    - name: Nelle Varoquaux
      affiliation: Department of Statistics and Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA 
    - name: Elizabeth Purdom
      affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA

>>>>

---
title: "[Normalization] A pipeline to analyse time-course gene expression data"
author:
    - name: Nelle Varoquaux
      affiliation: CNRS, University Grenoble Alpes, TIMC-IMAG, Grenoble, France 
    - name: Elizabeth Purdom
      affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
